Processing genome scale tabular data with wormtable

### 2013-08-04 docs/paper/paper.tex
Added discussion of interval search + minor edits

Another problem with tables stored text files is that it is difficult 
to index the information within particular columns. 
This means that most operations on a table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

The obvious solution to these problems is to load tabular data into a
relational database. Databases store values in binary form so that parsing is 
not required, and support efficient retrieval and indexing. There 
are many problems with this approach, however. Relational databases are 
extremely complex systems, each supporting different features and SQL 
dialects. It is not a straightforward task to design a schema for a 
particular data set, particularly not if portability across different 
databases is required. Similarly, accessing the data requires a knowledge
of SQL. In the most common case, a database server must also be maintained
and so storage and user permissions must be carefully managed. All of 
these aspects require significant expertise.

@@

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which dictate 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf}
a Python VCF parser.  In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes;
the wormtable version required
57 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 80 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that although Unix 
pipelines are efficient, it is a difficult and error prone method 
of processing data when more complex calculations are required.

@@

\subsection*{Seek performance}
One of the most serious problems with tabular data stored in text files
is that accessing a particular row efficiently without some auxiliary 
index. Without some information on where a particular row is located
in a file, there is nothing that can be done except to read the file 
line-by-line until the required row is found. Tabix~\cite{li11} solves
this problem by compressing a tab-delimited text file into blocks, and 
then storing an index mapping genome position to the location 
of the compressed block, and the position of the row within the block.
Tabix is specifically designed for range queries, allowing us to 
efficiently retrieve all of the rows within a given genomic range.

To compare the seek performance of wormtable with Tabix, we compressed
the VCF file mentioned in the previous subsection using \texttt{bgzip}
and indexed it using \texttt{tabix}. Since the VCF file was not 
strictly in order of genome position (in violation of the VCF format),
we first needed to sort the VCF file, which required 5m13s. Afterwards,
compression required 11m8s, and indexing with \texttt{tabix} required
1m12s. The size of the resulting compressed file was 2.9GB and 
the size of the index file was 106KB.
To duplicate the functionality of Tabix in wormtable, we simply need to 
create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
This required approximately
2 minutes and the resulting index consumed 115.4MB of space.

@@

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevent pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, compression is a double-edged sword: 
it does lead to much smaller files, which reduces the number of random seeks 
required.  On the other hand, though, decompression is expensive and must 
be repeated each time a block is accessed, unless some sophisticated
caching mechanism is devised.

@@

Wormtable is currently limited to supporting Python only, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

@@

We have concentrated on the VCF format here since it is a widely 
used and popular format. After converting the VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the columns of 
interest. For accessing regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. In the worst case, when using a 
slow hard disk this is somewhat slower than Tabix. In the best
case, when frequently accessing the same regions or using 
fast solid state storage, wormtable is many times faster than 
Tabix. Moreover, wormtable is much more general than Tabix, 
since it can index any column or combination of columns.
Similarly, wormtable is not limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.

>>>>

Another problem with tables stored as text files is that it is difficult 
to index the information within particular columns. 
This means that many operations on a table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

The obvious solution to these problems is to load tabular data into a
relational database. Databases store values in binary form so that parsing is 
not required, and support efficient retrieval and indexing. There 
are many problems with this approach, however. Relational databases are 
extremely complex systems, each supporting different features and SQL 
dialects. It is not a straightforward task to design a schema for a 
particular data set, particularly not if portability across different 
databases is required. Similarly, accessing data requires a knowledge
of SQL. In the most common case, a database server must also be maintained
and so storage and user permissions must be carefully managed. All of 
these aspects require significant expertise.

@@

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf}
a Python VCF parser.  In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes, whereas 
the wormtable version required
57 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 80 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that although Unix 
pipelines are efficient, it is a difficult and error prone method 
of processing data when more complex calculations are required.

@@

\subsection*{Seek performance}
One of the most serious problems with tabular data stored in text files
is that it is not possible to access a particular row 
efficiently without some auxiliary 
index. Unless we have some information on where a particular row is located
in a file, there is little that can be done except to read the file 
line-by-line until the required row is found. Tabix~\cite{li11} solves
this problem by compressing a tab-delimited text file into blocks, and 
then storing an index mapping genome position to the location 
of the compressed block, and the position of the row within the block.
Tabix is specifically designed for range queries, allowing us to 
efficiently retrieve all of the rows within a given genomic range.

To compare the seek performance of wormtable with Tabix, we compressed
the VCF file mentioned in the previous subsection using \texttt{bgzip}
and indexed it with \texttt{tabix}. Since the VCF file was not 
strictly in order of genome position (in violation of the VCF format),
we first needed to sort the VCF file, which required 5m13s. Afterwards,
compression required 11m8s, and indexing with \texttt{tabix} required
1m12s. The size of the resulting compressed file was 2.9GB and 
the size of the index file was 106KB.
To duplicate the functionality of Tabix on VCF data, we simply need to 
create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
This required approximately
2 minutes and the resulting index consumed 115MB of space.

@@

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevent pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, compression is a double-edged sword: 
it does lead to much smaller files, which reduces the number of random seeks 
required.  On the other hand, however, decompression is expensive and must 
be repeated each time a block is accessed unless some sophisticated
caching mechanism is devised.

@@

Wormtable is currently limited to supporting Python only, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. 
Wormtable does not support interval search, and so it is not
straightforward to find, for example, all rows overlapping 
a given genomic region in GTF files. This problem has been 
solved several times, however, and we aim to adapt existing 
techniques~\cite{li11,kzbhk10} and incorporate them into 
wormtable.  Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

@@

We have concentrated on the VCF format here since it is a widely 
used and popular format. After converting the VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the columns of 
interest. For accessing regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using the  \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. In the worst case, when using a 
slow hard disk this is somewhat slower than Tabix. In the best
case, when frequently accessing the same regions or using 
fast solid state storage, wormtable is many times faster than 
Tabix. Moreover, wormtable is much more general than Tabix, 
since it can index any column or combination of columns.
Similarly, wormtable is not limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.
